### 2018-06-03 doc/manuscript.Rmd
todo to done, first draft of revision (not including citations)

There are some possible explanations to these findings.

{{
There are two sources for endogenous fatty acids: diet and de novo lipogenesis 
(DNL). Given that most fatty acids can come from either source, it is difficult
to

}}

Previous research has shown that carbohydrate intake increases de novo lipogenesis (DNL)
[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. 
Several studies have shown a
link between higher estimated DNL and an increased risk for metabolic 
dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a]; however, how
DNL may influence metabolic dysfunction is not well described. One possible explanation
may be the relationship between greater DNL and the types of fatty acids produced.

The link between higher DNL and increases in specific fatty acids has been
examined in several studies. Previous studies that have examined DNL have used
markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or 16:1n-7
to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are 
limitations to using these ratios as the fatty acids used in their calculation
can also be obtained from the diet in addition to being created through DNL
[@Hodson2008a]. A feeding trial (n=24) was conducted to identify 
the fatty acids that most accurately reflected DNL as potential biomarkers
[@Lee2015a]. The study found that palmitoleic acid (16:1n-7), directly measured
DNL using isotopes, and liver fat were all highly correlated with each other
(r>0.50), suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In
another small (n=14) feeding trial, meal type (high fat vs low fat) was tested
to determine its effect on DNL and TGFA composition [@Wilke2009a]. The authors
reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. These fatty acids are similar to the fatty acids we found
that clustered together using the PLS analysis, implicating these fatty acids as
indicative of a higher carbohydrate diet. A higher carbohydrate diet, 
particularly one characterized by a predominance of simple carbohydrates, may
lead to greater DNL in an attempt to control blood glucose, thus increasing hepatic
fat stores and consequently increasing the amount of TGFA in circulation
[@Chehade2013a; @Verges2015a]. We found that
higher proportions of these four DNL TGFA also highly correlated with a higher
concentration of clinical TG, reinforcing this pathway between carbohydrate
intake, DNL, and circulating TG. The higher concentration of circulating 14 and
16 carbon fatty acids may then expose tissues to greater lipotoxicity, for
instance from palmitic acid (16:0), which is well-known to have harmful effects
on tissues [@Riserus2008a; @Iggman2010a].

Our study 
extends these findings by showing that TGFA with 14 to 16 carbons clustered
together and this pattern strongly predicted lower insulin sensitivity. While
these fatty acids also had a significant association with beta-cell function,
the magnitude of associations were more modest compared to those for insulin
sensitivity.

The direction of association between TGFA and insulin sensitivity is unclear 
from previous cross-sectional studies due to the physiological feedback mechanisms 
involved. For example, while greater DNL may promote muscle insulin resistance, 
the reverse may also be true [@Flannery2012a]. Higher insulin resistance may
encourage greater DNL to handle the higher blood glucose. {{higher insulin
resistance may also lead to higher circulating LDL, since they are not taken
up... cite}}
As we found no interaction by time of TGFA on insulin sensitivity, this study
can not determine the exact role of the feedback mechanism. Combined with the
lack of time interaction and the consistent negative association without the
time interaction, these results at least suggest that the feedback mechanism may
not be strongly influential and that TGFA may predict insulin sensitivity over a
six year period. Given the complex biological mechanisms and feedback loops
involved, disentangling whether insulin sensitivity influences TGFA more
strongly than TGFA influencing insulin sensitivity will require more
complex research designs and analyses.

Given the close biological relationship between circulating NEFA and TG, NEFA 
may act as confounding factor and was thus adjusted for. In a previous published
analysis of the NEFA fraction [@Johnston2017b], we found that higher total NEFA,
but not the specific composition, associated with lower beta-cell function. This
is in contrast to the TGFA findings that the specific composition does
differentially associate with insulin sensitivity and beta-cell function,
adjusting for total NEFA. There was no difference in results in models that did
not include NEFA as a confounder (data not shown). This difference in results
between NEFA and TGFA suggests that TGFA may independently and strongly influence the 
pathophysiology of T2D, when compared to other lipid fraction compositions,
including the phospholipid and cholesteryl ester fractions [@Johnston2016a]. This
may be because TG are biologically destined for uptake by non-hepatic tissue as
they are found mainly in VLDL, at least during fasting. This is in contrast to
NEFA that are mostly uptaken by the liver and used in TG production [@Nielsen2012a].

<!--
{{ comment on comparing models with and without NEFA }} While NEFA may be a
mediating factor, we found that adjustment for it did not attenuate this
association since TGFA composition is also dictated by other sources for DNL
(e.g. recent meal, DNL from glucose stores, fatty acid recycling, etc.).

- contributions to DNL (e.g. 70% 16:0 to newly synth) {{ confirm what this is }}

>>>>

There are a few possible explanations to these findings. Fatty acids, and thus
the TGFA composition, derive from two sources: dietary fat and de novo
lipogenesis (DNL). Adipose tissue also supplies circulating fatty acids,
however, these also ultimately originate from the two sources. Specific dietary
fat and carbohydrates can influence DNL activity {{cite}}. Determining the
specific source of fatty acids is extremely difficult to ascertain outside of
highly controlled experimental settings. However, there are sign that suggest
one source over another.

The link DNL activity and levels of specific fatty acids has been examined in
several studies. Previous studies that have examined DNL have used
markers of estimated DNL, such as the ratio between 18:2n-6 to 16:0 or 16:1n-7
to 16:0 [@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are 
limitations to using these ratios as the fatty acids used in their calculation
can also be obtained from the diet in addition to being created through DNL
[@Hodson2008a]. A feeding trial (n=24) was conducted to identify 
the fatty acids that most accurately reflected DNL as potential biomarkers
[@Lee2015a]. The study found that palmitoleic acid (16:1n-7), directly measured
DNL using isotopes, and liver fat were all highly correlated with each other
(r>0.50), suggesting that 16:1n-7 may be a good biomarker for hepatic DNL. In
another small (n=14) feeding trial, meal type (high fat vs low fat) was tested
to determine its effect on DNL and TGFA composition [@Wilke2009a]. The authors
reported that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. {{talk about Yki-.. study}}.  In our findings, the four
fatty acids were highly correlated amongst each other, in addition to clustering
together on their negative association with insulin sensitivity. If these fatty
acids derived from mainly the diet, we would anticipate seeing more correlation
with the 18:0, 18:1, and 18:2 fatty acids, as these fatty acids often share a
dietary source (e.g. meat) {{cite?}} with the four clustering fatty acids. As our
results do not show this, this suggests the common source being DNL and that the
fatty acids may have been produced from higher DNL activity since, again, there
was no strong correlation with 18:0 and 18:1 fatty acids, which would be the
next stage DNL via elongation and/or desaturation of 16:0 and 16:1 fatty acids.

A higher DNL activity may result from a higher carbohydrate intake as a means to
control blood glucose [@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a;
@Parks1999a; @Chehade2013a; @Verges2015a], which could increase hepatic fat
stores and increase the amount of TG in circulation. In fact, we found that
higher proportions of these four TGFA also highly correlated with a higher
concentration of clinical TG. Several studies have shown a link between higher
estimated DNL and an increased risk for metabolic dysfunction [@Ma2015a;
@Zong2013a; @Lankinen2015a; @Kroger2011a]; how DNL may influence metabolic
dysfunction is not well understood. Possible explanations may be that higher DNL
produces more of certain fatty acids or that higher DNL increases circulating TG,
which itself is well documented to contribute to metabolic dysfunction {{cite?}}.

Regardless of the exact source of these fatty acids, our results in addition to
the available scientific evidence emphasizes the importance of the fatty acid
composition on metabolic health, as individual fatty acids can have specific
physiological functions {{cite}}. For instance, a higher concentration of
circulating 14 and 16 carbon fatty acids may expose tissues to greater
lipotoxicity, for instance from palmitic acid (16:0), which is well-known to
have harmful effects on tissues [@Riserus2008a; @Iggman2010a]. Our study extends
these findings by showing that TGFA with 14 to 16 carbons clustered together and
this pattern strongly predicted lower insulin sensitivity. While these fatty
acids also had a significant association with beta-cell function, the magnitude
of associations were more modest compared to those for insulin sensitivity.

The direction of association between TGFA and insulin sensitivity is unclear
from previous cross-sectional studies due to the physiological feedback
mechanisms involved. For example, while greater DNL and higher TGFA may promote
muscle insulin resistance, the reverse may also be true [@Flannery2012a]. Higher
insulin resistance may encourage TGFA to remain in circulation longer {{cite?}}.
As we found no interaction by time of TGFA on insulin sensitivity, this study
cannot determine the exact role of the feedback mechanism. However, combining the
with the lack of time interaction and the consistent negative association
without the time interaction, these results at least suggest that the feedback
mechanism may not be strongly influential and that TGFA may predict insulin
sensitivity over a six year period. Given the complex biological mechanisms and
feedback loops involved, disentangling whether insulin sensitivity influences
TGFA more strongly than TGFA influencing insulin sensitivity will require more
complex research designs and analyses.

Given the close biological relationship between circulating NEFA and TG, NEFA 
may act as confounding factor and was thus adjusted for. In our published
analysis of the NEFA fraction [@Johnston2017b], we found that higher total NEFA,
but not the specific composition, associated with lower beta-cell function. This
is in contrast to the TGFA findings that the specific composition does
differentially associate with insulin sensitivity and beta-cell function,
adjusting for total NEFA. There was no difference in results in models that did
not include NEFA as a confounder (data not shown). This difference in results
between NEFA and TGFA suggests that TGFA may independently and strongly influence the 
pathophysiology of T2D, when compared to other lipid fraction compositions,
including the phospholipid and cholesteryl ester fractions [@Johnston2016a]. This
may be because TG are biologically destined for uptake by non-hepatic tissue as
they are found mainly in VLDL, at least during fasting. This is in contrast to
NEFA that are mostly uptaken by the liver and used in TG production [@Nielsen2012a].

<!--
- contributions to DNL (e.g. 70% 16:0 to newly synth) {{ confirm what this is }}
